---
title: "Archetypes"
author: "James Thorson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Archetypes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Description

We here show how to replicate and extract the archetypes estimated in Thorson et al. 2023.  We first load packages and data:

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# Build
#  setwd(R'(C:\Users\James.Thorson\Desktop\Git\FishLife)'); devtools::build_rmd("vignettes/Archetypes.Rmd"); rmarkdown::render( "vignettes/Archetypes.Rmd", rmarkdown::pdf_document())
```

```{r, echo=TRUE, message=FALSE}
# Load packages and data
library(archetypes)
library(FishLife)
data(FishBase_and_Morphometrics)

# Set seed so that archetypes are labeled the same way when replicating
suppressWarnings(RNGversion("3.5.0"))
set.seed(1986)
```

# Replicate model fitting

If we want to replicate the entire analysis, we can load and inspect the SEM specification:

```{r, echo=TRUE, message=FALSE, fig.width=6, fig.height=6, eval=TRUE}
#
text = FishBase_and_Morphometrics$text
cat(text)
```

We also need the trait values, relatedness structure, and identify association of trait columns with factor levels.  To match the analysis in the main text (using taxonomy to represent relatedness), we specify `Z_ik` rather than `tree`:
```{r, echo=TRUE, message=FALSE, fig.width=6, fig.height=6, eval=TRUE}
# load tree
Y_ij = FishBase_and_Morphometrics$Y_ij

# load taxonomy
Z_ik = FishBase_and_Morphometrics$Z_ik

# Identify number of levels for categorical traits
n_traits = 21
n_levels = c( rep(1,17), 2, 4, 2, 4)
group_j = rep(seq_len(n_traits)-1, times=n_levels)
```

We then can run this using the loaded data.  Here, we start it from the previous maximum-likelihood estimate to save time:

```{r, echo=TRUE, message=FALSE, fig.width=6, fig.height=6, eval=FALSE}
#
startpar = unlist(FishBase_and_Morphometrics$ParHat[c("L_z","obsL_z")])
startpar = startpar[which(startpar!=0.01)]

# Full run
Estimate_database = Fit_model(
  text = text,
  Database = NULL,
  Y_ij = Y_ij,
  Z_ik = Z_ik,
  min_replicate_measurements = 3,
  startpar = startpar,
  group_j = group_j,
  multinomial_for_factors = TRUE,
  extract_covariance = FALSE,
  getsd = FALSE,
  newtonsteps = 0,
  add_predictive = FALSE )
```

# Replicate archetype analysis

Alternative, we might want to replicate the archetype analysis and plot the resulting simplex of archetypes.  To do so, we use the recorded estimates to save time:

```{r, echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
# Load results
beta_iv = FishBase_and_Morphometrics$beta_gv[ FishBase_and_Morphometrics$g_i, ]
a <- archetypes(beta_iv, 3, verbose = TRUE)

# Labels
Labels = c("Equil.","Oppor.","Per.")[c(1,3,2)]
Col_names = colnames(FishBase_and_Morphometrics$Y_ij)
  Col_names = gsub("habitat", "H: ", Col_names )
  Col_names = gsub("body_shape", "BS: ", Col_names )
  Col_names = gsub("feeding_mode", "FM: ", Col_names )
  Col_names = gsub("spawning_type", "ST: ", Col_names )

# Col function
mycol = function(x,alpha){
  y = col2rgb(x)
  rgb( y[1,1]/255, y[2,1]/255, y[3,1]/255, alpha=alpha)
}

# Plot by taxa
Z_ik = FishBase_and_Morphometrics$Z_ik
simplex = simplexplot(a, show_points=FALSE, labels=Labels) #, projection=atypes_projection)
points( simplex$proj_h[which(Z_ik$Family=="Gobiidae"), ], col=mycol("brown",0.5), pch=20 )
points( simplex$proj_h[which(Z_ik$Family=="Percidae"), ], col=mycol("orange",0.5), pch=20 )
points( simplex$proj_h[which(Z_ik$Family=="Salmonidae"), ], col=mycol("red",0.5), pch=20 )
points( simplex$proj_h[which(Z_ik$Family=="Scorpaenidae"), ], col=mycol("grey",0.5), pch=20 )
points( simplex$proj_h[which(Z_ik$Family=="Clupeidae"), ], col=mycol("green",0.5), pch=20 )
points( simplex$proj_h[which(Z_ik$Class=="Elasmobranchii"), ], col=mycol("black",0.5), pch=20 )
points( simplex$proj_h[which(Z_ik$Family=="Poeciliidae"), ], col=mycol("blue",0.5), pch=20 )
points( simplex$proj_h[which(Z_ik$Family=="Atherinidae"), ], col=mycol("purple",0.5), pch=20 )
legend( "topleft", bty="n", ncol=2, legend=c("Gobiidae","Salmonidae","Poeciliidae",
        "Atherinidae","Percidae","Scorpaenidae","Clupeidae","Elasmobranchii"),
        fill=c("brown","red","blue","purple","orange","grey","green","black"))
```

If interested, we can pull out values for individual species based on their Latin binomial.  We here show a few examples:

```{r, echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
# Pull out examples
Z_ik = FishBase_and_Morphometrics$Z_ik
species_to_plot = data.frame(
  c("Gadus_macrocephalus", "Hippoglossus_stenolepis", "Sebastes_alutus", "Thaleichthys_pacificus",
     "Oncorhynchus_tshawytscha", "Somniosus_pacificus", "Mallotus_villosus", "Boreogadus_saida",
     "Bothrocara_brunneum", "Tarletonbeania_taylori", "Gambusia_affinis", "Callogobius_sclateri",
     "Acanthurus_xanthopterus", "Eviota_jewettae"),
  c( "Pacific cod", "Halibut", "Paciric Ocean perch", "Eulachon", "Chinook", "Sleeper shark",
     "Capelin", "Arctic cod", "twoline eelpout", "North Pacific lanternfish",
     "Western mosquitofish", "Pacific goby", "Yellowfin surgeonfish", "Jewett's dwarfgoby")
)
species_to_plot$color = rainbow(nrow(species_to_plot))
# Plot by taxa
simplex = simplexplot(a, show_points=FALSE, labels=Labels)
for( r in seq_len(nrow(species_to_plot)) ){
  which_row = which(paste0(Z_ik$Genus,"_",Z_ik$Species)==species_to_plot[r,1])
  text( x=simplex$proj_h[which_row,1], y=simplex$proj_h[which_row,2],
        col=species_to_plot[r,3], labels=letters[r] )
}
legend( "topleft", bty="n", ncol=2, legend=species_to_plot[,2], fill=species_to_plot[,3])
```

Finally, we can plot the distribution of predicted traits for all species, along with the predicted mean for each archetype:

```{r, echo=TRUE, message=FALSE}
Col_names = colnames(FishBase_and_Morphometrics$beta_gv)
  Col_names = gsub("habitat", "H: ", Col_names )
  Col_names = gsub("body_shape", "BS: ", Col_names )
  Col_names = gsub("feeding_mode", "FM: ", Col_names )
  Col_names = gsub("spawning_type", "ST: ", Col_names )
dims = c(6,6)
par( mfrow=dims, mgp=c(2,0.5,0), mar=c(1,1,2,0), tck=-0.02 )
for( cI in 1:ncol(FishBase_and_Morphometrics$beta_gv) ){
  hist( FishBase_and_Morphometrics$beta_gv[,cI],
        main=Col_names[cI], xlab="", ylab="", yaxt="n" )
  abline( v=parameters(a)[,cI], col=viridisLite::viridis(3), lwd=3 )
}
```